% !TEX encoding = ISO-8859-16
\section{Radon Wasserstein Barycenters}
\label{sec-bary-radon}

Proposition~\ref{prop-bary-1d} shows that it is computationally inexpensive to compute the Wasserstein barycenter of 1-D densities. It thus makes sense to seek for alternate definitions of barycenters of measures in $\RR^d$ that rely on 1-D Wasserstein distances and barycenters. This section investigates a construction based on the Radon transform.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Radon Transform of Functions}

We recall below classical definitions, and refer to~\cite{Helgason-radonbook} for more details. The Radon transform is first defined on integrable functions. 

\begin{defn}[Radon transform of functions]
The Radon transform $Rf$ of $f \in \Lun(\RR^d)$ is defined as
\eql{\label{eq-radon}
		Rf(t,\th) = \int_{\RR^{d-1} } f(t\theta + U_\th \ga) \d \ga
}
where $U_\th \in \R^{d\times(d-1)}$ is any matrix such that its  columns defines an orthogonal basis of $\th^\perp$ (the hyperplane orthogonal to $\th$). 
%  such that $\Theta = [\th, U_\th] \in \R^{d\times d}$ is an orthonormal basis of $\R^d$, {i.e.} the columns of $U_\th$ span the hyperplane $\Hh_{0,\th^{\perp}}$ which passes through the origin and is normal to $\th$.  
This defines $R : \Lun(\RR^d) \rightarrow \Lun(\Om^d)$. 
\end{defn}


\if 0
%% OLD DEF %%%
\eq{
	\foralls (t,\th) \in \Om^d, \quad  
	Rf(t,\th) =
	\int_{ \Hh_{t,\th} } f \d m_{\Hh_{t,\th}} 
	%= {\color{red} \int_{\ga \in \R^{d-1} } f(t\theta + U_\th \ga) d\ga}
	% \int_{ \dotp{x}{\th}=t } f(x) \d x 
}
\eq{
	\qwhereq
	\Hh_{t,\th} = \enscond{x \in \RR^d}{ \dotp{x}{\th}=t }
}
and $m_{\Hh_{t,\th}}$ is the Lebesgues measure on the hyperplane $\Hh_{t,\th}$.
\fi


Its adjoint is defined on continuous functions as follows.

\begin{defn}[Back-projection operator] 
The back projection $R^*g$ of $g \in \Cont{\Om^d}$ is defined as
\eq{
	R^* g(x) % = \int_{ \dotp{x}{\th}=t } g(t,\th) \d t \d \th
	 = \int_{ \Sph } g(\dotp{x}{\th},\th) \d \th.
}
This defines $R^* : \Cont{\Om^d} \rightarrow \Cont{\RR^d}$. 
\end{defn}

\newcommand{\filt}{h}

One has that $R^* R$ is a translation invariant operator, i.e. a convolution  
\eq{
	R^*R f = \filt \star f
	\qwhereq
	\hat\filt(\om) = c \, \norm{\om}^{-(d-1)},
}
where $\star$ is the convolution on $\RR^d$ and $c \in \RR$ is a normalizing constant whose exact value depends on the dimension (see~\cite{Helgason-radonbook}). This relationship suggests a definition of a pseudo-inverse transform which operates on smooth functions so as to invert the low pass filter $\filt$.

\begin{defn}[Inverse Radon transform of functions] 
The pseudo-inverse Radon transform $R^+ g$ of $g \in \Dd(\Om^d)$ is defined as
\eql{\label{eq-pseudo-inv}
	R^{+} g = \filt^+ \star (R^* g)
}
where $\filt^+$ is defined through $\hat \filt^+(\om) = c^{-1} \norm{\om}^{d-1}$.
\end{defn}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Radon Transform of Measures}

Since $R^*$ is defined on $\Cont{\RR^d}$, the Radon transform is naturally extended to measures $\mu \in \Mm(\RR^d)$ by duality as follows.

\begin{defn}[Radon transform of measures]\label{defn-radon-measure} For all $\mu \in \Mm(\RR^d)$, we set $\nu=R(\mu)$ be defined through, $\foralls g \in \Cont{\Om^d}$, 
\eql{\label{eq-radon-measure} 
	\int_{\Om^d} g(t,\th) \d\nu(t,\th)
	= \int_{\RR^d} (R^*g)(x) \d \mu(x).
} 
This defines $R : \Mm(\RR^d) \rightarrow \Mm(\Om^d)$.
\end{defn}

The following proposition shows that the Radon transform of a measure gathers projections of the input measure along all possible directions.

\begin{prop}\label{prop-radon-pushforward}
For $\mu \in \Mm(\RR^d)$, one has 
\eq{
	\foralls \th \in \Sph, \quad
	R(\mu)^{\th} = P_\th \sharp \mu
}
\eq{
	\qwhereq
	P_\th : x \in \RR^d \mapsto \dotp{x}{\th} \in \RR,
}
and where $R(\mu)^{\th} \in \Mm(\RR)$ is defined in~\eqref{eq-desintegration}.
\end{prop}

The proof of this proposition, as well as all the other proofs of this section, can be found in Appendix~\ref{sec-appendix-radon}.


The conditional measure $\nu^\th$ associated to $\nu \in \Mm(\Om^d)$ is defined for almost all $\th$, i.e. on a Borel set of $\th \in \Sph$ of measure 1. Proposition~\ref{prop-radon-pushforward} shows that when $\nu=R(\mu)$, then $\nu^\th$ is in fact well defined for all $\th \in \Sph$, because it is a push-forward measure. 


\if 0
Since $R$ is defined on $\Lun(\RR^d)$ and thus on $\Cont(\RR^d)$, it is possible to define $R^*$ on measures by duality. 

\begin{defn}[Adjoint Radon transform of measures]\label{defn-adj-radon-measure} For all $\nu \in \Mm(\Om^d)$, we set $\mu=R^*(\nu)$ be defined through, $\foralls f \in \Cont{\RR^d}$, 
\eql{\label{eq-radon-measure} 
	\int_{\Om^d} g(t,\th) \d\nu(t,\th)
	= \int_{\RR^d} (R^*g)(x) \d \mu(x).
} 
This defines $R : \Mm(\RR^d) \rightarrow \Mm(\Om^d)$.
\end{defn}
\fi


We define in a way similar to Definition~\ref{defn-radon-measure} the inverse Radon transform using the operator $R^{+,*} = R (R^*R)^{-1}$.

\begin{defn}[Inverse Radon transform of measures] For all $\nu \in \Mm(\Om^d)$, we set $\mu=R^+(\nu) \in \Dd^*(\RR^d)$ be defined through, $\foralls f \in \Dd(\RR^d)$, 
\eql{\label{eq-inv-radon-measure}
	\int_{\RR^d} f(x) \d\mu(x)
	= \int_{\Om^d} (R^{+,*}f)(t,\th) \d \nu(t,\th).
}
This defines $R^+ : \Mm(\Om^d) \rightarrow \Dd^*(\RR^d)$. 
\end{defn}

Note that for an arbitrary $\nu \in \Mm(\Om^d)$ (i.e. not necessarily in the range $\Im(R)$ of $R$), $R^+\nu$ is a distribution and not necessarily a measure. One can however show that for $\nu = R(\mu) \in \Im(R)$, then $R^+(\nu)=\mu \in \Mm(\RR^d)$ is a measure, as detailed in the following proposition. The proof of this proposition can be found in~\cite{Boman-Radon}, Section~3.
%  for more details and connections with the celebrated Cram\`er-Wold Theorem.


\begin{prop}\label{prop-im-radon}
	$R : \Mm(\RR^d) \rightarrow \Mm(\Om^d)$ defined in~\eqref{eq-radon-measure} is injective, and $R^+R=\Id_{\Mm(\RR^d)}$.
\end{prop}


The following lemma recapitulates useful commutation properties of the Radon transform with respect to translation and scaling.

\begin{lem}\label{lem-invariances}
		One has, for $\mu \in \Mm_1^+(\RR^d)$ and $\nu \in \Mm_1^+(\Om^d)$, and for all 
		$(s,u,\Phi) \in \RR^{+,*} \times \RR^d \times \Oo(\RR^d)$, 
		\begin{align}
		  	\label{propR} R(\phi_{s,u}\sharp\mu) &= \psi_{s,u}\sharp R(\mu)  \\ 
%		  	\label{propRm} R^*(\psi_{s,u}\sharp\nu) &=  s^{d-1} \cdot (\phi_{s,u}\sharp R^*(\nu)) \\
		  	\label{propRp} R^+(\psi_{s,u}\sharp\nu) &=  \phi_{s,u}\sharp R^+(\nu) \\
			\label{PropRot} R( \Phi \sharp \mu ) &= \tilde\Phi \sharp R(\mu).
		\end{align}			
\end{lem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Radon Barycenter}

According to Proposition~\ref{prop-im-radon}, one has 
\eq{
	R : \Mm_1^+(\RR^d) \rightarrow R(\Mm_1^+(\RR^d)) \subset \bar\Mm_1^+(\Om^d),  
}
although the inclusion on the right hand side is not an equality. This property allows us to define the Radon barycenter.

\begin{defn}[Radon barycenter]\label{defn-radon-baryc} Given $\la \in \La_I$ and $(\mu_i)_{i \in I} \in \Mm_1^+(\RR^d)^I$, we define
\eq{
	\Bary{\RR^d}^R(\mu_i,\la_i)_{i \in I} = 
	R^+ \Bary{\Om^d}^W(R(\mu_i),\la_i)_{i \in I} \in \Dd^*(\RR^d).
}
\end{defn}

Since for $\nu \in \Bary{\Om^d}^W(R(\mu_i),\la_i)_{i \in I}$ one does not have in general $\nu \in \Im(R)$, $\Bary{\RR^d}^R(\mu_i,\la_i)_{i \in I}$ is composed of distributions and not necessarily measures.


The following proposition shows that the Radon barycenter enjoys the same invariance properties to scaling, translation and rotation as the classical Wasserstein barycenter. 


\begin{prop}
\label{prop:InvarianceHolds}
	Proposition~\ref{prop-invariance-W} holds when replacing $\Bary{\RR^d}^W$ by $\Bary{\RR^d}^R$.
\end{prop}



The following proposition shows that, similarly to the usual Wassertstein barycenter, the Radon barycenter of translated and scaled copies of a given measure is also a translated and scaled copy.

\begin{prop}
\label{prop:InvarianceHoldsBis}
	Proposition~\ref{prop-invariance-W-bis} holds when replacing $\Bary{\RR^d}^W$ by $\Bary{\RR^d}^R$.
\end{prop}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Approximate Computation with Eulerian Discretization}
\label{subsec-algorithm-eulerian}

\input{sections/sec-bary-radon-discretization}
